Investigation of thermal reactionomes reveals differing responses to thermal

### 2014-10-01 doc/ApTranscriptome_manuscript.Rmd
updates to ms

---
title: Investigation of thermal reactionomes reveals differing responses to thermal
  extremes in warm and cold-climate ant species
author: "null"
csl: ecology.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document: null
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1}, James Vincent \textsuperscript{2},
Mahesh Vangala \textsuperscript{2}, Robert R. Dunn \textsuperscript{3}, Nathan J. Sanders \textsuperscript{4},
Nicholas J. Gotelli \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1}

@@

\textsuperscript{1} Department of Biology, University of Vermont \newline
\textsuperscript{2} Vermont Genetics Network, University of Vermont \newline
\textsuperscript{3} Department of Biological Sciences, North Carolina State University \newline
\textsuperscript{4} Center for Macroecology, Evolution and Climate, University of Copenhagen

@@

Thermal tolerance is a key trait governing the distributions of species and underlies their future responses to climate change. Here, we characterize the thermal “reactionome” – the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient to understand the genetic basis of thermal tolerance. We characterized the reactionomes of two common ant species in eastern U.S. forests: the northern cold-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*. We found that at least 10% of genes were thermally responsive. The warm-adapted *A. carolinensis* displayed higher basal levels of gene expression, upregulated fewer heat shock proteins, and exhibited greater inertia of gene expression chang than did *A. picea*. This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability. *A. carolinensis* also shut down more genes in response to thermal extremes than *A. picea*. A subset of these genes were upregulated in *A. picea*, including gene ontology terms such as metabolism, cellular organization, glycoprotein synthesis and DNA repair. These differences between species suggest that the two closely related species employ different strategies for coping with temperature stress. Namely *A. picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas *A. carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress. More broadly, this study adds to a growing body of evidence that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.

@@

Successfully coping with progressive changes in climate will require both short-term physiological acclimation and longer-term adaptive changes. To better understand the relationship between these two timescales, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A. carolinensis*. We found surprisingly divergent transcriptomic responses in the two species. *A. picea* responded to high and low temperatures through upregulation of a larger proportion of its transcriptome with a more robust heat shock response. In contrast, *A. carolinensis* reacted to thermal stress to a greater extent through downregulation, reducing metabolic gene expression and requiring more extreme temperatures to trigger a protective response. These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely-encountered stressors, which may limit adaptive potential of northern species in the face of rapidly shifting climatic zones.

@@

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. Outside of candidate genes [@krebs1998], little is known about the genetic basis of this variation or how it has evolved. To address this, we need information on the extent to which geographic differentiation and selection has acted upon the diversity of genes, and their plasticity of expression, involved in thermal tolerance. This information is a critical for understanding the adaptive potential of species to future climates [@hoffmann2008] and the link between short-term physiological acclimation and longer-term adaptation to increased temperatures [@somero2010].

@@

Regardless of the experimental design or treatment of temperature as a categorical variable with few levels, an important suite of issues often hampers many RNAseq studies that are designed to identify the genetic basis of environmental adaptation [@meyer2011; @teets2012; @barshis2013; @oneil2014]. Namely, lists of differentially-expressed genes derived from RNAseq results are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], may miss genes with small expression changes that have detectable phenotypic effects [@yan2002; @padovan2013], identify genes that are correlated, but not causative, of phenotypes [@fay2004] and may not be directly informative about organismal status [@feder2005]. Genome-wide association studies have faced a similar problem: phenotypic effects of a small number of statistically-significant genes are over-estimated while small effects of a large number of other causal genes are missed [@beavis1994; @larsson2013]. Our understanding of the genetic basis of traits, and our ability to predict evolutionary potential, are compromised by these issues.

@@

Second, we used the thermal reactionomes to examine the relationship between short-term physiological acclimation and longer-term adaptive changes to temperature. Specifically, through the process of genetic assimilation [@pigliucci2006], adaptive responses are hypothesized to involve the same molecular pathways as short-term acclimatory responses. Finding that the genes up-regulated in response to cold temperatures in *A. carolinensis* have increased basal expression in the colder-climate species *A. picea*, and vice versa, would be consistent with this prediction. While none of the above predictions depend on the functional annotation of the responsive genes [@pavlidis2012], we used gene set enrichment analysis to gain insight on the molecular toolkit involved in adaptation to thermal extremes. For example, in response to thermal extremes, we expect genes in the cellular stress response [@kultz2005] to be up-regulated, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated [@somero2011].

To test the above predictions for these two exemplar species, we collected whole organism gene expression data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina short read sequencing technology, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.

@@

The combined *Aphaenogaster* transcriptome assembly contained 99,861 transcripts. Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts (hereafter referred to as genes) with expression patterns that were thermally-responsive at 5% False Discovery Rate. Of these, `prettyNum(length(quadratic.lms), big.mark = ",")` had a non-linear relationship with temperature that would likely have been missed with a standard differential expression approach. In effect, `r round(9100/99861, 2)*100`% of the transcriptome is thermally-responsive. This result is the same if we focus on only the transcripts that have a BLAST hit. However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts (i.e. the closest related ant species with sequenced genomes have only about 17,000 genes [@gadau2012]) indicating that this estimate is a lower-bound for the true proportion of genes that are thermally-responsive.

@@

pandoc.table(Ap.type.table[,c("Low", "Intermediate", "High", "Bimodal", "NotResp")], style="rmarkdown",
             big.mark = ",", split.tables = Inf, 
             caption = "Table of the number of thermally-responsive genes by expression type for *A. carolinensis* and *A. picea*. **Low** are genes with increased expression at low temperatuers (< 10°C), **Intermediate** are genes with maximum expression between 10 - 30°C, **High** are genes with increased expression at high temperatures (> 31°C), **Bimodal** are genes with increased expression at both low and high temperatures, while **NotResp** are genes that are thermally-responsive in one species but not the other.")
```

We found that the molecular diversity of thermal response to cold temperatures is substantially greater than of that to hot temperatures. In particular, we grouped the thermally-responsive genes for each species into five expression categories based on their reaction norms: genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (**Not Responsive**). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C). In both species,  the majority of genes were in the **Low** category (Table 1). For *A. carolinensis*, this was followed by **Intermediate**, **High** and then **Bimodal**, whereas for *A. picea* the second largest category was **Bimodal** followed by **High** and finally **Intermediate** (Table 1).

Gene set enrichment analysis (GSEA) of the responsive genes indicated that the molecular responses to thermal extremes involve large-scale cellular and organizational changes, though the molecular responses to high temperatures are more limited than those to low temperatures. Specifically, out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` terms enriched (*P* < 0.01) for **Low** genes. These terms grouped into 15 clusters, including "protein modification", "protein metabolic process", "lipid metabolic process", "regulation of signaling", "protein transport/localization", "detection of stimulus", and "chondroitin sulfate biosynthesis". For *High* genes, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched terms, with 5 of the 7 terms being involved in "regulation of biosynthetic/metabolic process".  For **Bimodal** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched and grouped into 5 clusters, including "biological regulation", "regulation of cellular biosynthesis", "gene expression", "signaling", and "protein modification process". For **Intermediate** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched and grouped into 3 clusters including "lipid biosynthesis/metabolism", "glycoprotein biosynthesis/metabolism", and "DNA organization".

# Shifts in the expression patterns of thermally-responsive genes between species

@@

In contrast to our prediction that more genes would be thermally-responsive in the warmer-climate species *A. carolinensis*, the total number of thermally-responsive genes did not differ between the species (...r t.test(Ap.type.table[1,1:4], Ap.type.table[2,1:4])...), but differed significantly in how they were distributed among the expression categories (Stuart-Maxwell test of marginal homogeneity ... ). Specifically, compared to a hypothetical homogenous distribution, there were an excess of **Intermediate** genes in *A. carolinensis* which disproportionatly switched to **Low** or **Bimodal** expression categories in *A. picea* (Fig. 1).

The molecular processes in this set of genes that switched from down-regulation in *A. carolinensis* to up-regulation in *A. picea* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` GO terms, of which ... were related to metabolism. In addition, there were GO terms related to repair (e.g. "histone modification"), gene regulation (e.g. "positive regulation of cellular process") and cellular organization (e.g. "transport", "organelle organization"). These results indicate that in response to thermal extremes, the warmer-climate species, *A. carolinensis*, shuts down metabolism while the colder-climate species, *A. picea*, increases metabolism.

@@

# posthoc result for each cell
Ec.cells <- sign(tt2 - Ec) * (tt2 - Ec)^2 / Ec
Ec.cells.sign.table <- apply(Ec.cells, c(1,2), function(x) ifelse(abs(x) <= X2.df, "0", 
                                                            ifelse(x > X2.df, "+", "-")))

md <- melt(Ec.cells)

@@

To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the mean temperatures of gene activation, defined as the temperatures at which the gene showed the greatest change in expression, for genes turned on at high or low temperatures, including *Bimodal* genes. Genes turned on at high temperatures in *A. carolinensis* were activated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A. picea* (*t*~df=`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For genes with increased expression at low temperatures, there was a negligible difference in the mean temperature of gene activation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~df=`r round(T_on.low.ttest$parameter, 0)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2).

@@

By using a regression-approach, we are able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes. We found that about 70% of the thermally-responsive genes (7,202) have non-linear responses to temperature.

This result is consistent with the only other work we are aware of that examines reaction norms of gene expression, which found that most genes in yeast have have expression differences in response to copper exposure [@hodgins-davis2012]. In general, finding thousands of genes that respond to temperature highlights the limitations of previous methods that only found a few dozen to hundred genes are involved in thermal response from recent studies in corals [@meyer2011; @barshis2013; @palumbi2014] and butterflies [@oneil2014] which only used a few temperature conditions. Underlying the importance of thermal tolerance as a physiological trait [@kingsolver2009], the genetic basis of thermal tolerance is likely to be much more complex than indicated by these previous studies.

@@

```{r max_exp_PDF, eval = FALSE, echo = FALSE, warning = FALSE, message = FALSE, fig.cap = "Probability density function of the temperature of maximum expression for all thermally-responsive genes in each species."}
# reshape data
Ap.df$Species <- ifelse(Ap.df$colony == "ApVT", "A. picea", "A. carolinensis")

maxexplot <- ggplot(Ap.df, aes(x=max.val, fill=Species)) + 
  geom_density(alpha=0.2, position="identity") + 
  labs(x = "Temperature of maximum expression", y = "Density")

print(maxexplot)

####################################
### CHANGE TO REACTION NORM FOR HSPs 
####################################

```

@@

Thermal safety margins...

>>>>

---
title: Investigation of thermal reactionomes reveals differing responses to thermal
  extremes in warm and cold-climate ant species
author: 'null'
csl: ecology.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document: null
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1}, James Vincent \textsuperscript{2},
Mahesh Vangala \textsuperscript{2}, Robert R. Dunn \textsuperscript{3}, Aaron M. Ellison \textsuperscript{4}, Nathan J. Sanders \textsuperscript{5}, Nicholas J. Gotelli \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1}

@@

\textsuperscript{1} Department of Biology, University of Vermont, Burlington, Vermont 05405 \newline
\textsuperscript{2} Vermont Genetics Network, University of Vermont, Burlington, Vermont 05405 \newline
\textsuperscript{3} Department of Biological Sciences, North Carolina State University, Raleigh, North Carolina 27695 \newline
\textsuperscript{4} Harvard Forest, Harvard University, Petersham, Massachusetts 01336 \newline
\textsuperscript{5} Center for Macroecology, Evolution and Climate, University of Copenhagen

@@

Thermal tolerance is a key trait governing the distributions of species and underlies their future responses to climate change. However, little is known about the genetic basis of thermal tolerance in most species. Here, we describe the thermal “reactionome”, the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient, to understand the genetic basis of thermal tolerance. We characterized the reactionomes of two common ant species in eastern U.S. forests: the northern cold-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*. We found that at least 10% of genes were thermally responsive. The warm-adapted *A. carolinensis* displayed higher basal levels of gene expression, upregulated fewer heat shock proteins, and exhibited greater inertia of gene expression change than did *A. picea*. This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability. *A. carolinensis* also shut down more genes in response to thermal extremes than *A. picea*. A subset of these genes were upregulated in *A. picea*, including gene ontology terms such as metabolism, cellular organization, glycoprotein synthesis and DNA repair. These differences between species suggest that the two closely related species employ different strategies for coping with temperature stress. Namely *A. picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas *A. carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress. More broadly, this study adds to a growing body of evidence that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.

@@

Successfully coping with progressive changes in climate will require both short-term physiological acclimation and longer-term adaptive changes. To better understand the relationship between these two timescales, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A. carolinensis*. We found surprisingly divergent transcriptomic responses in the two species. *A. picea* responded to high and low temperatures through upregulation of a larger proportion of its transcriptome with a more robust heat shock response. In contrast, *A. carolinensis* reacted to thermal stress to a greater extent through downregulation, reducing metabolic gene expression and requiring more extreme temperatures to trigger a protective response. These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely-encountered stressors, which may limit adaptive potential of the northern species in the face of rapidly shifting climatic zones.

@@

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. Outside of candidate genes [@krebs1998], little is known about the genetic basis of this variation or how it has evolved. To address this, we need information on the extent to which geographic differentiation and selection has acted upon the diversity of genes, and their plasticity of expression, involved in thermal tolerance. This information is a critical for understanding the adaptive potential of species to future climates [@hoffmann2008] and the link between short-term physiological acclimation and longer-term adaptation to increased temperatures [@somero2011].

@@

Regardless of the experimental design or treatment of temperature as a categorical variable with few levels, a suite of issues hampers many RNAseq studies that are designed to identify the genetic basis of environmental adaptation [@meyer2011; @teets2012; @barshis2013; @oneil2014]. Namely, lists of differentially-expressed genes derived from RNAseq results are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], may miss genes with small expression changes that have detectable phenotypic effects [@yan2002; @padovan2013], identify genes that are correlated, but not causative, of phenotypes [@fay2004] and may not be directly informative about organismal status [@feder2005]. Genome-wide association studies have faced a similar problem: phenotypic effects of a small number of statistically-significant genes are over-estimated while small effects of a large number of other causal genes are missed [@beavis1994; @larsson2013]. Our understanding of the genetic basis of traits, and our ability to predict evolutionary potential, are compromised by these issues.

@@

Second, we used the thermal reactionomes to examine the relationship between short-term physiological acclimation and longer-term adaptive changes to temperature. Specifically, through the process of genetic assimilation [@pigliucci2006], adaptive responses are hypothesized to involve the same molecular pathways as short-term acclimatory responses. Finding that the genes up-regulated in response to cold temperatures in *A. carolinensis* have increased basal expression in the colder-climate species *A. picea*, and vice versa, would be consistent with this prediction. C

While none of the above predictions depend on the functional annotation of the responsive genes [@pavlidis2012], we used gene set enrichment analysis to gain insight on the molecular toolkit involved in adaptation to thermal extremes. For example, in response to thermal extremes, we expect genes in the cellular stress response [@kultz2005] to be up-regulated, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated [@somero2011]. To test the above predictions for these two exemplar species, we collected whole organism gene expression data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina short read sequencing technology, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.

@@

The combined *Aphaenogaster* transcriptome assembly contained 99,861 transcripts. Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts (hereafter referred to as genes), `r round(9100/99861, 2)*100`% of the transcriptome, with expression patterns that were thermally-responsive at 5% False Discovery Rate. Of these, `r prettyNum(length(quadratic.lms), big.mark = ",")` had a non-linear relationship with temperature that would likely have been missed with a standard differential expression approach. However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts (i.e. the closest related ant species with sequenced genomes have only about 17,000 genes [@gadau2012]) indicating that this estimate is a lower-bound for the true proportion of genes that are thermally-responsive.

@@

pandoc.table(Ap.type.table[,c("Low", "Intermediate", "High", "Bimodal", "NotResp")], style="rmarkdown",
             big.mark = ",", split.tables = Inf, 
             caption = "Number of thermally-responsive genes by expression type for *A. carolinensis* and *A. picea*. **Low** are genes with increased expression at low temperatuers (< 10°C), **Intermediate** are genes with maximum expression between 10 - 30°C, **High** are genes with increased expression at high temperatures (> 31°C), **Bimodal** are genes with increased expression at both low and high temperatures, while **NotResp** are genes that are thermally-responsive in one species but not the other.")
```

We found that the number of genes that are responsive to cold temperatures is substantially greater than of that to hot temperatures (Table 1). In particular, we grouped the thermally-responsive genes for each species into five expression categories based on their reaction norms: genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (**Not Responsive**). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C). In both species,  the majority of genes were in the **Low** category (Table 1). For *A. carolinensis*, this was followed by **Intermediate**, **High** and then **Bimodal**, whereas for *A. picea* the second largest category was **Bimodal** followed by **High** and finally **Intermediate** (Table 1).

Gene set enrichment analysis (GSEA) of the responsive genes indicated that the molecular responses to thermal extremes involve large-scale cellular and organizational changes. Specifically, out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` terms enriched (*P* < 0.01) for **Low** genes. These terms grouped into 15 clusters, including "protein modification", "protein metabolic process", "lipid metabolic process", "regulation of signaling", "protein transport/localization", "detection of stimulus", and "chondroitin sulfate biosynthesis". For *High* genes, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched terms, with 5 of the 7 terms being involved in "regulation of biosynthetic/metabolic process".  For **Bimodal** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched and grouped into 5 clusters, including "biological regulation", "regulation of cellular biosynthesis", "gene expression", "signaling", and "protein modification process". For **Intermediate** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched and grouped into 3 clusters including "lipid biosynthesis/metabolism", "glycoprotein biosynthesis/metabolism", and "DNA organization".

## Shifts in the expression patterns of thermally-responsive genes between species

@@

```{r, echo=FALSE}
Ap.t.test <- t.test(Ap.type.table[1,1:4], Ap.type.table[2,1:4])
```

In contrast to our prediction that more genes would be thermally-responsive in the warmer-climate species *A. carolinensis*, the total number of thermally-responsive genes did not differ between the species (*t*~df~=`r round(Ap.t.test$parameter, 0)` = `r round(Ap.t.test$statistic, 2)`, *P* = `r round(Ap.t.test$p.value, 2)`), but differed significantly in how they were distributed among the expression categories (Stuart-Maxwell test of marginal homogeneity ... ). Specifically, compared to a hypothetical homogenous distribution, there were an excess of **Intermediate** genes in *A. carolinensis* which disproportionatly switched to **Low** or **Bimodal** expression categories in *A. picea* (Fig. 1).

The molecular processes in this set of genes that switched from down-regulation in *A. carolinensis* to up-regulation in *A. picea* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` GO terms, of which ... were related to metabolism. In addition, there were GO terms related to repair (e.g. "histone modification"), gene regulation (e.g. "positive regulation of cellular process") and cellular organization (e.g. "transport", "organelle organization"). These results suggest that in response to thermal extremes, the warmer-climate species *A. carolinensis* shuts down metabolism while the colder-climate species *A. picea* increases metabolism, potentially to support cellular resistance to thermal stress.

@@

# posthoc result for each cell
Ec.cells <- sign(tt2 - Ec) * (tt2 - Ec)^2 / Ec
Ec.cells.sign.table <- apply(Ec.cells, c(1,2), function(x) ifelse(abs(x) <= X2.df, "0", 
                                                            ifelse(x > X2.df, "+", "-")))
md <- melt(Ec.cells)

@@

To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the mean temperatures of gene activation, defined as the temperatures at which the gene showed the greatest change in expression, for genes turned on at high or low temperatures, including *Bimodal* genes. Genes turned on at high temperatures in *A. carolinensis* were activated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A. picea* (*t*~df=`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For genes with increased expression at low temperatures, there was a negligible difference in the mean temperature of gene activation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~df~=`r round(T_on.low.ttest$parameter, 0)` = `r round(T_on.low.ttest$statistic, 1)`, *P* = `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2).

@@

Species persistence in face of climate change will require short-term physiological acclimation as well as longer-term adaptive changes. In this study, we explored the relationship between these processes using whole organism gene expression data from closely-related ant species. By applying a regression-approach to gene expression data, we were able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes. In the combined transcriptome of these species, we found that a minimum of 9% of genes are thermally-responsive, and that about 70% of these genes (7,202) have non-linear responses to temperature that would not have been detected by standard differential expression approaches.

The result that a large fraction of the genome is responsive is consistent with the only other work we are aware of that examines reaction norms of gene expression, which found that most genes in yeast have have expression differences in response to copper exposure [@hodgins-davis2012]. In general, finding thousands of genes that respond to temperature highlights the limitations of previous methods that only found a few dozen to hundred genes are involved in thermal response from recent studies in corals [@meyer2011; @barshis2013; @palumbi2014] and butterflies [@oneil2014] which only used a few temperature conditions. Underlying the importance of thermal tolerance as a physiological trait [@kingsolver2009], the genetic basis of thermal tolerance is likely to be much more complex than indicated by these previous studies.

@@

@@
